Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-10 background.tex
+

A relatively under-explored space of modal logics is the interaction of modal and substructural logics, 
in particular hybrid-style modalities in substructural logics, which has seen only minimal exploration~\todo{copy refs from fs paper} and no prior 
application. We develop our ideas in the Iris program logic~\cite{jung2018iris}, as a result exploring additional subtleties 
that arise where the modality itself may entail ownership of resources, as well as interactions between our hybrid-style 
modality and substructural rules.  For example, Iris contains a number of modalities that distribute over separating 
conjunction, or for which resources can freely move into the modality 
(e.g., $\blacktriangleright(P)\ast Q \wand \blacktriangleright(P\ast Q)$). In our setting some of these rules 
apply while others do not. For example, in our setting, an assertion that involves no modalities is interpreted as 
holding in the current (active-on-the-CPU) address space, so clearly cannot move into arbitrary other address spaces 
--- unless guarded by another address space modality.

>>>>

A relatively under-explored space of modal logics is the interaction of modal and substructural logics, 
in particular hybrid-style modalities in substructural logics, which has seen only minimal exploration~\cite{dovsen1992modal,restall1993modalities,d1997grafting,kamide2002kripke,licata2017fibrational} and no prior 
application. We develop our ideas in the Iris program logic~\cite{jung2018iris}, as a result exploring additional subtleties 
that arise where the modality itself may entail ownership of resources, as well as interactions between our hybrid-style 
modality and substructural rules.  For example, Iris contains a number of modalities that distribute over separating 
conjunction, or for which resources can freely move into the modality 
(e.g., $\blacktriangleright(P)\ast Q \wand \blacktriangleright(P\ast Q)$). In our setting some of these rules 
apply while others do not. For example, in our setting, an assertion that involves no modalities is interpreted as 
holding in the current (active-on-the-CPU) address space, so clearly cannot move into arbitrary other address spaces 
--- unless guarded by another address space modality.
%
%
%Work on modal logic extends back for decades with many widely-varied variations; we discuss here only the pieces most relevant to our work.
%Broadly speaking modal logics incorporate \emph{modal operators}, which take as arguments a proposition expected to be true in another time~\cite{pnueli1977temporal}, place~\cite{gordon2019modal}\todo{other cites}, or circumstance~\cite{hintikka1962knowledge,halpern1985guide}, and result in a proposition true in the \emph{current} time, place, or circumstance at which the truth of the use of the modal operator is being evaluated. Classic examples include modal necessity $\square P$ describing that $P$ is \emph{necessarily} true, $\mathsf{G}~P$ meaning $P$ is true \emph{globally} (i.e., forever from this time onwards), or $K_i(P)$ describing that a particular participant $i$ \emph{knows} that $P$ is true. The latter is an example of of \emph{multimodal} logic, where there is an indexed family of modalities (modal operators) parameterized by some dimension of interest (there, participants).
%A closely related variant of multimodal logic is \emph{dynamic logic}~\cite{pratt1976semantical}, a logic of weakest preconditions~\cite{dijkstra-75} which works with modalities of the form $[p](P)$, which states that \emph{in the current program state}, \emph{if} program $p$ is run then afterwards $P$ will hold (modulo non-termination).
%This same idea is used to encode Hoare triples in the Iris program logic~\cite{krebbers2017essence}, using the same encoding as in Pratt's original presentation, where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$. Unlike classic work on dynamic logic~\cite{harel2000dynamic}, Iris applies these ideas in a \emph{substructural} setting (separation logic) where distributivity laws over substructural connectives must be considered. For both the dynamic modality and the later modality $\blacktriangleright$ used for guarded recursive predicates, these modalities satisfy axioms of the form $M(P)\ast Q\vdash M(P\ast Q)$, but not the reverse.
%Iris is not the first combination of modal and substructural logic~\cite{dovsen1992modal,restall1993modalities,d1997grafting,kamide2002kripke,licata2017fibrational}, but is certainly the most heavily tested.
%A hallmark of a unary logical operator $M$ being a modality is if $M$ satisfies an property akin to $(\upvarphi\rightarrow\psi)\rightarrow M(\upvarphi)\rightarrow\psi$, which roughly states that modus ponens holds under the modality.\footnote{Afficionados of modal logic will note that this property is not quite Axiom K (which requires the initial implication to also hold under $M$), but follows from K and a necessitation rule $\upvarphi\rightarrow M(\upvarphi)$. Non-necessitive modalities typically satisfy the weaker property we call out above.}
%Another pillar of our work is hybrid logic, a branch of modal logic with \emph{names} (called \emph{nominals} for states in Kripke models~\cite{blackburn1995hybrid,goranko1996hierarchies,areces2001hybrid,gargov1993modal}, in contrast to the typical flavor of modality which refers to an \emph{unspecified} (in the assertion) set of alternative circumstances. We are aware of only two prior works combining hybrid logics with program logics specifically.  Brotherston and Villard~\cite{brotherston2014parametric} demonstrated that may properties true of various separation logics are not definable in boolean \BI (\BBI), and showed that a hybrid extension \HyBBI allows most such properties to be defined (e.g., the fact that separating conjunction is cancellative is unprovable in boolean \BI, but provable in \HyBBI). There, nominals named resources (roughly, but not exactly, heap fragments). Gordon~\cite{gordon2019modal} described a use of hybrid logic in the verification of actor programs, where nominals named the local state of individual actors (with such assertions stabilized with a rely/guarantee approach). Beyond these, there is limited work on the interaction of specifically \emph{hybrid logic} with substructural logics. Primarily there is a line of work on hybrid linear logic (\HyLL)~\cite{despeyroux2014hybrid}, originally used as a way to more conveniently express aspects of transition systems in linear logic. However, \HyLL's proof rules offer no non-trivial interactions with multiplicative connectives (every \HyLL proof can in fact be embedded into regular linear logic~\cite{chaudhuri2019hybrid}, unlike Brotherston and Villard's \HyBBI, which demonstrably increases expressive power over its base \BBI.\colinsays{some of this is too detailed for this point in the paper, needs to go to related work. But I'll split it later.}
%
%In both \HyLL and \HyBBI, nominals denote worlds with monadic structure (as worlds in Kripke semantics for either LL or \BBI necessarily have monoidal structure). Our nominals, by contrast, do not name worlds in the same sense with respect to Iris's CMRAs, but in fact \emph{classes} of worlds, because the names are locations (a means of \emph{selecting} resources) rather than resources.  A key difference is that the use of nominals in those logics corresponds specifically to hypothetical reasoning about resources (until a nominal is connected to a current resource, in which case conclusions can be drawn about the current resource), which means the modalities themselves do not ``own'' resources. Instead, our hybrid modalities directly own portions of the data structure...\colinsays{also need to move to related work}
%
%Our work draws on ideas from hybrid modal logics, applied in the context of Iris's higher-order separation logic~\cite{jung2018iris}. We define our modalities and prove their rules directly within Iris, so there is no model-theoretic novelty in our work. Instead, we show that targeted use of modalities combining the ideas of \emph{named resources} with the power of substructural reasoning enable clear specification of general reasoning principles for data structures. Along the way we take advantage of the fact that modalities need not be injections between the \emph{same} logics, but in fact the propositions that modal operators contextualize can come from \emph{other} logics; in our case we exploit the fact that Iris itself is parameterized by a choice of step-indexed~\cite{ahmed-appel-virga-02} \textsf{BI}-algebra~\cite{ohearn1999bunched} to allow convenient specifications in embedded data-structure-relative logics.
